---
title: "The Health and Economic Consequences of Storms"
author: "Matthew Hayes"
output: html_document
---

## Synopsis
Tornados resulted in the most fatalities, both in total and on average. Most injuries come from tornados (total) and heat waves (on average). Property and crop damage share a common theme: too much water isn't good. Floods, hurricanes, and typhoons top the lists, both in total and on average. When considering combined health consequences, tornados result in the most harm by far, but on average one should be more concerned with heat waves, tropical storms, and wildfires. As for economic consequences, it can't be said enough: Water and wind are costly.

## Data Processing

The original dataset, [found here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2), is very expansive and we won't need all of it for this analysis. A little pre-processing will make the data much more workable.

First our packages:

```{r prepack}
library(dplyr)
library(tidyr)
```

It's important to note that `NA` values in this dataset are represented by empty strings.

```{r read, cache = TRUE}
storm <- tbl_df(read.csv("./storm.csv.bz2", stringsAsFactors = FALSE, 
                         na.strings = c(NA, "")
))
storm
```

We'll only need seven variables: `EVTYPE`, `FATALITIES`, `INJURIES`, `PROPDMG`, `PROPDMGEXP`, `CROPDMG`, and `CROPDMGEXP`.

```{r subset, cache = TRUE}
storm <- storm %>%
    select(EVTYPE, FATALITIES:CROPDMGEXP)
names(storm) <- tolower(names(storm))
storm
```

According to the [storm data documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), the variables `propdmgexp` and `cropdmgexp` are multipliers, where K = thousands, M = millions, and B = billions. Do they take any other values in our data set?

```{r unique}
unique(storm$propdmgexp)
unique(storm$cropdmgexp)
```

Looks like we have some data entry issues. Given this, we need to remove any records that do not match our known multipliers.

```{r multremove}
storm <- mutate(storm, propdmgexp = tolower(propdmgexp),
                cropdmgexp = tolower(propdmgexp))
good <- c("k", "m", "b", NA) 
gProp <- storm$propdmgexp %in% good
gCrop <- storm$cropdmgexp %in% good
storm$propdmgexp[!gProp] <- NA
storm$cropdmgexp[!gCrop] <- NA
rm(good, gProp, gCrop)
unique(storm$propdmgexp)
unique(storm$cropdmgexp)
```

Let's convert these multipliers to their respective numerical equivalents and use them to find the true values of property and crop damage.

```{r mult}
storm <- storm %>%
    mutate(propdmgexp = gsub("k", 1000, propdmgexp),
           propdmgexp = gsub("m", 1000000, propdmgexp),
           propdmgexp = gsub("b", 1000000000, propdmgexp),
           cropdmgexp = gsub("k", 1000, cropdmgexp),
           cropdmgexp = gsub("m", 1000000, cropdmgexp),
           cropdmgexp = gsub("b", 1000000000, cropdmgexp),
           propdmgexp = as.numeric(propdmgexp),
           cropdmgexp = as.numeric(cropdmgexp),
           propdmg = propdmg * propdmgexp,
           cropdmg = cropdmg * cropdmgexp
    ) %>%
    select(-propdmgexp, -cropdmgexp)
storm
```

I prefer working with tall datasets. Let's tidy the dataset by gathering fatalities and injuries, as well as property and crop damage.

```{r gather}
storm <- storm %>%
    gather(dmgType, dmgCount, -(evtype:injuries)) %>%
    gather(injType, injCount, -evtype, -(dmgType:dmgCount))
storm
```

## Health Consequences

Across the United States, which types of events are most harmful with respect to population health? For this analysis, we'll use `fatalities` and `injuries` to represent health consequences.

We'll need some extra packages.

```{r packhealth}
library(ggplot2)
library(gridExtra)
```


Using our `storm` data, let's look at total and mean fatalities/injuries by event type.

```{r health}
health <- storm %>%
    group_by(evtype, injType) %>%
    summarize(total = sum(injCount, na.rm = TRUE),
              avg = mean(injCount, na.rm = TRUE)
    ) %>%
    ungroup()
health
```

What are the top three "killer" storms, in total and on average?

```{r healthfat}
fatTotal <- health %>%
    filter(injType == "fatalities") %>%
    arrange(desc(total)) %>%
    head(3)
fatTotal
fatAvg <- health %>%
    filter(injType == "fatalities") %>%
    arrange(desc(avg)) %>%
    head(3)
fatAvg
```

Tornados are by far the deadliest storms, results in `r fatTotal[1, 3]` deaths. It also kills the most per storm. Do tornados also injure the most people?

```{r healthink}
injTotal<- health %>%
    filter(injType == "injuries") %>%
    arrange(desc(total))  %>%
    head(3)
injTotal
injAvg <- health %>%
    filter(injType == "injuries") %>%
    arrange(desc(avg)) %>%
    head(3)
injAvg
```

While tornados have also injured by far the most people, it appears that heat waves, on average, produce the most injuries.

Let's visualize these findings.

```{r healthplot, fig.height = 8, fig.width = 12}
h1 <- ggplot(fatTotal, aes(x = reorder(evtype, -total), y = total)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Total Fatalities", title = "Highest Fatalities")
h2 <- ggplot(fatAvg, aes(x = reorder(evtype, -avg), y = avg)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Mean Fatalities",
         title = "Average Fatalities per Event")
h3 <- ggplot(injTotal, aes(x = reorder(evtype, -total), y = total)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Total Injuries", title = "Highest Injuries")
h4 <- ggplot(injAvg, aes(x = reorder(evtype, -avg), y = avg)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Mean Injuries", 
         title = "Average Injuries per Event")
grid.arrange(arrangeGrob(h1, h2, h3, h4, main = "Effects on Population Health"))
```

## Economic Consequences

Across the United States, which types of events have the greatest economic consequences? We'll use `propdmg` and `cropdmg` as proxies for economic damage.

As we did with `health`, let's look at total and mean property/crop damage by event type.

```{r econ}
econ <- storm %>%
    group_by(evtype, dmgType) %>%
    summarize(total = sum(dmgCount, na.rm = TRUE),
              avg = mean(dmgCount, na.rm = TRUE)
    ) %>%
    ungroup()
econ
```

Which storms are the most costly with respect to property?

```{r property}
propTotal <- econ %>%
    filter(dmgType == "propdmg") %>%
    arrange(desc(total)) %>%
    head(3)
propTotal
propAvg <- econ %>%
    filter(dmgType == "propdmg") %>%
    arrange(desc(avg)) %>%
    head(3)
propAvg
```

Taken loosely, too much water and wind is a very bad thing when it comes to property. Does this hold true for crop damage as well?

```{r crop}
cropTotal<- econ %>%
    filter(dmgType == "cropdmg") %>%
    arrange(desc(total))  %>%
    head(3)
cropTotal
cropAvg <- econ %>%
    filter(dmgType == "cropdmg") %>%
    arrange(desc(avg)) %>%
    head(3)
cropAvg
```

Most definitely. Unlike health consequences, economic consequences tend to be consistent across property and crops, as well as in total and on average.

Let's visualize our results.

```{r econplot, fig.height = 8, fig.width = 16}
e1 <- ggplot(propTotal, aes(x = reorder(evtype, -total), y = total)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Total (dollars)", 
         title = "Highest Property Damage")
e2 <- ggplot(propAvg, aes(x = reorder(evtype, -avg), y = avg)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Mean (dollars)",
         title = "Average Property Damage per Event")
e3 <- ggplot(cropTotal, aes(x = reorder(evtype, -total), y = total)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Total (dollars)",
         title = "Highest Crop Damage")
e4 <- ggplot(cropAvg, aes(x = reorder(evtype, -avg), y = avg)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Mean (dollars)",
         title = "Average Crop Damage per Event")
grid.arrange(arrangeGrob(e1, e2, e3, e4, main = "Economic Consequences"))
```

## Results

Across the United States, which types of events are more costly with respect to human health and economic well-being? We'll combine `fatalities` and `injuries` to represent human health. Economic well-being will be represented by combining `propdmg` and `cropdmg`.

First let's combine our data.

```{r combo}
healthCombo <- health %>%
    group_by(evtype) %>%
    summarize(total = sum(total),
              avg = mean(avg)) %>%
    ungroup()
healthCombo
econCombo <- econ %>%
    group_by(evtype) %>%
    summarize(total = sum(total),
              avg = mean(avg)) %>%
    ungroup()
econCombo
```

Next, we'll look at the top three offenders in each category.

```{r comborank}
healthComboTotal <-healthCombo %>%
    arrange(desc(total)) %>%
    head(3)
healthComboTotal
healthComboAvg <-healthCombo %>%
    arrange(desc(avg)) %>%
    head(3)
healthComboAvg
econComboTotal <- econCombo %>%
    arrange(desc(total)) %>%
    head(3)
econComboTotal
econComboAvg <- econCombo %>%
    arrange(desc(avg)) %>%
    head(3)
econComboAvg
```

Tornados result in the most harm by far, but on average one should be more concerned with heat waves, tropical storms, and wildfires.

As for economic consequences, it can't be said enough: Water and wind are costly.

Finally, let's visualize it.

```{r comboplot, fig.height = 7, fig.width = 14}
c1 <- ggplot(healthComboTotal, aes(x = reorder(evtype, -total), y = total)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Total Fatalities/Injuries", 
         title = "Highest Human Health Consequences")
c2 <- ggplot(healthComboAvg, aes(x = reorder(evtype, -avg), y = avg)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Mean Fatalities/Injuries",
         title = "Average Human Health Consequences per Event")
c3 <- ggplot(econComboTotal, aes(x = reorder(evtype, -total), y = total)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Total Damage (dollars)",
         title = "Highest Economic Damage")
c4 <- ggplot(econComboAvg, aes(x = reorder(evtype, -avg), y = avg)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Mean Damage (dollars)",
         title = "Average Economic Damage per Event")
grid.arrange(arrangeGrob(c1, c2, c3, c4, 
                         main = "Health and Economic Consequences"))
```