---
title: "The Health and Economic Consequences of Storms"
author: "Matthew Hayes"
date: "12/20/2014"
output: html_document
---

## Synopsis
While tornados resulted in the most fatalities, both in total and on average, most injuries come from wildfires (total) and heat waves (on average). Property and crop damage share a common theme: too much water isn't good. Floods, hurricanes, and typhoons top the lists, both in total and on average.

## Data Processing

The original dataset, [found here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2), is very expansive and we won't need all of it for analysis. A little pre-processing will make the data more workable.

First our packages:

```{r prepack}
library(dplyr)
library(tidyr)
```

Assuming you've downloaded the data, we need to read it in. It's important to note that `NA` values are represented by empty strings.
```{r read, cache = TRUE}
storm <- tbl_df(read.csv("./storm.csv.bz2", stringsAsFactors = FALSE, 
                         na.strings = c(NA, "")
))
storm
```

We'll only need seven variables: `EVTYPE`, `FATALITIES`, `INJURIES`, `PROPDMG`, `PROPDMGEXP`, `CROPDMG`, and `CROPDMGEXP`.

```{r subset, cache = TRUE}
storm <- storm %>%
    select(EVTYPE, FATALITIES:CROPDMGEXP)
names(storm) <- tolower(names(storm))
storm
```

The variables `propdmgexp` and `cropdmgexp` are multipliers, where K = thousands, M = millions, and B = billions. Do they take any other values in our data set?

```{r unique}
unique(storm$propdmgexp)
unique(storm$cropdmgexp)
```

Looks like we have some data entry issues. Given this, we need to remove any records that do not match our known multipliers.

```{r multremove}
storm <- mutate(storm, propdmgexp = tolower(propdmgexp),
                cropdmgexp = tolower(propdmgexp))
good <- c("k", "m", "b", NA) 
gProp <- storm$propdmgexp %in% good
gCrop <- storm$cropdmgexp %in% good
storm$propdmgexp[!gProp] <- NA
storm$cropdmgexp[!gCrop] <- NA
rm(good, gProp, gCrop)
unique(storm$propdmgexp)
unique(storm$cropdmgexp)
```

Let's convert these multipliers to their respective numerical equivalents and use them to find the true values of property and crop damage.
```{r mult}
storm <- storm %>%
    mutate(propdmgexp = gsub("k", 1000, propdmgexp),
           propdmgexp = gsub("m", 1000000, propdmgexp),
           propdmgexp = gsub("b", 1000000000, propdmgexp),
           cropdmgexp = gsub("k", 1000, cropdmgexp),
           cropdmgexp = gsub("m", 1000000, cropdmgexp),
           cropdmgexp = gsub("b", 1000000000, cropdmgexp),
           propdmgexp = as.numeric(propdmgexp),
           cropdmgexp = as.numeric(cropdmgexp),
           propdmg = propdmg * propdmgexp,
           cropdmg = cropdmg * cropdmgexp
    ) %>%
    select(-propdmgexp, -cropdmgexp)
storm
```

I prefer working with tall datasets. Let's tidy the dataset by gathering fatalities and injuries, as well as property and crop damage.

```{r gather}
storm <- storm %>%
    gather(dmgType, dmgCount, -(evtype:injuries)) %>%
    gather(injType, injCount, -evtype, -(dmgType:dmgCount))
storm
```

## Health Consequences

Across the United States, which types of events are most harmful with respect to population health? For this analysis, we'll use `fatalities` and `injuries` to represent population health consequences.

We'll need some extra packages.
```{r packhealth}
library(ggplot2)
library(gridExtra)
```


Using our `storm` data, let's look at total and mean fatalities/injuries by event type.

```{r health}
health <- storm %>%
    group_by(evtype, injType) %>%
    summarize(total = sum(injCount, na.rm = TRUE),
              avg = mean(injCount, na.rm = TRUE)
    ) %>%
    ungroup()
health
```

What are the top three "killer" storms, in total and on average?

```{r healthfat}
fatTotal <- health %>%
    filter(injType == "fatalities") %>%
    arrange(desc(total)) %>%
    head(3)
fatTotal
fatAvg <- health %>%
    filter(injType == "fatalities") %>%
    arrange(desc(avg)) %>%
    head(3)
fatAvg
```

Tornados are by far the deadliest storms, results in `r fatTotal[3, 1]` deaths. It also kills the most per storm. Do tornados also injure the most people?

```{r healthink}
injTotal<- health %>%
    filter(injType == "injuries") %>%
    arrange(desc(total))  %>%
    head(3)
injTotal
injAvg <- health %>%
    filter(injType == "injuries") %>%
    arrange(desc(avg)) %>%
    head(3)
injAvg
```

While tornados have also injured by far the most people, it appears that heat waves, on average, produce the most injuries.

Let's visualize these findings.

```{r healthplot, fig.height = 8, fig.width = 12}
h1 <- ggplot(fatTotal, aes(x = reorder(evtype, -total), y = total)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Total Fatalities", title = "Highest Fatalities")
h2 <- ggplot(fatAvg, aes(x = reorder(evtype, -avg), y = avg)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Mean Fatalities",
         title = "Average Fatalities per Event")
h3 <- ggplot(injAvg, aes(x = reorder(evtype, -total), y = total)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Total Injuries", title = "Highest Injuries")
h4 <- ggplot(injAvg, aes(x = reorder(evtype, -avg), y = avg)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Mean Injuries", 
         title = "Average Injuries per Event")
grid.arrange(arrangeGrob(h1, h2, h3, h4, main = "Effects on Population Health"))
```

## Economic Consequences

Across the United States, which types of events have the greatest economic consequences? We'll use `propdmg` and `cropdmg` as proxies for economic damage.

As we did with `health`, let's look at total and mean property/crop damage by event type.

```{r econ}
econ <- storm %>%
    group_by(evtype, dmgType) %>%
    summarize(total = sum(dmgCount, na.rm = TRUE),
              avg = mean(dmgCount, na.rm = TRUE)
    ) %>%
    ungroup()
econ
```

Which storms are the most costly with respect to property?

```{r property}
propTotal <- econ %>%
    filter(dmgType == "propdmg") %>%
    arrange(desc(total)) %>%
    head(3)
propTotal
propAvg <- econ %>%
    filter(dmgType == "propdmg") %>%
    arrange(desc(avg)) %>%
    head(3)
propAvg
```

Taken loosely, too much water and wind is a very bad thing when it comes to property. Does this hold true for crop damage as well?

```{r crop}
cropTotal<- econ %>%
    filter(dmgType == "cropdmg") %>%
    arrange(desc(total))  %>%
    head(3)
cropTotal
cropAvg <- econ %>%
    filter(dmgType == "cropdmg") %>%
    arrange(desc(avg)) %>%
    head(3)
cropAvg
```

Most definitely. Unlike health, economic consequences tend to be consistent across property and crops, as well as in total and on average.

Let's visualize our results.

```{r econplot, fig.height = 8, fig.width = 16}
e1 <- ggplot(propTotal, aes(x = reorder(evtype, -total), y = total)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Total (dollars)", 
         title = "Highest Property Damage")
e2 <- ggplot(propAvg, aes(x = reorder(evtype, -avg), y = avg)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Mean (dollars)",
         title = "Average Property Damage per Event")
e3 <- ggplot(cropTotal, aes(x = reorder(evtype, -total), y = total)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Total (dollars)",
         title = "Highest Crop Damage")
e4 <- ggplot(cropAvg, aes(x = reorder(evtype, -avg), y = avg)) +
    geom_bar(stat = "identity") +
    labs(x = "Event Type", y = "Mean (dollars)",
         title = "Average Crop Damage per Event")
grid.arrange(arrangeGrob(e1, e2, e3, e4, main = "Economic Consequences"))
```